---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# elm 

## Overview

`elm` is a package for import and type conversion of metadata statements. It's
function is to enable users to store metadata in markdown files, and convert
them to EML.

If you have any comments, questions or suggestions, please [contact us](mailto:support@ala.org.au).
 
## Installation
This package is under active development, and is not yet available on CRAN. You
can install the latest development version from GitHub with:

```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("atlasoflivingaustralia/elm")
```

Load the package:

```{r, eval = FALSE}
library(elm)
```


## Basic usage

The basic use case for `elm` is to convert markdown to eml in as few steps 
as possible:

```{r, eval=FALSE}
x <- read_md_chr("testfile.md")
x # intermediate object is a tibble
write_md_xml(x, "testout.xml")
```

## Citing `elm`
To generate a citation for the package version you are using, you can run:

```r
citation(package = "elm")
```

The current recommended citation is:

> Westgate MJ, Balasubramaniam S & Kellie D (2024) elm: Convert markdown files
to EML. R Package version 0.1.0.9999.