---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# elm 

## Overview

`elm` is a package for import and type conversion of metadata statements. It's
function is to enable users to store metadata in markdown files, and convert
them to EML.

If you have any comments, questions or suggestions, please [contact us](mailto:support@ala.org.au).
 
## Installation
This package is under active development, and is not yet available on CRAN. You
can install the latest development version from GitHub with:

```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("atlasoflivingaustralia/elm")
```

Load the package:

```{r, eval = FALSE}
library(elm)
```


## Features

`elm` contains tools to:

 - Import markdown files to R
 - convert to `tibble`s, `eml` (list) format, or `xml`


## Citing `elm`
To generate a citation for the package version you are using, you can run:

```r
citation(package = "elm")
```

The current recommended citation is:

> Westgate MJ, Balasubramaniam S & Kellie D (2024) elm: Convert markdown files
to EML. R Package version 0.1.0.9999.