---
title: "Quick start guide"
author: "Martin Westgate"
date: "2024-11-06"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The `paperbark` package converts markdown files to Ecological Metadata Language (EML),
which is a "comprehensive vocabulary and a readable XML markup syntax for 
documenting research data" 
([Jones et al. 2019](https://doi.org/10.5063/F11834T2)). EML has been widely
adopted as a community-maintained standard for metadata exchange, to the point 
that supplying metadata in this format is a requirement of several data-sharing
institutions, such as the Global Biodiversity Information Facility
([GBIF](https://gbif.org])) and its partner nodes. Unfortunately, however, the 
nested structure and xml formatting of EML make it difficult for people to use
without drawing on bespoke programming tools. In contrast, markdown is 
considerably more generic, especially for those who are familiar with tools such 
as RMarkdown or Quarto; but no general-purpose parser exists to transform 
markdown into EML. `paperbark` fills this gap.

## Methodological approach

To convert between markdown and EML, `paperbark` relies on the assumption that header
levels in markdown reflect the nested structure of the resulting xml. For
example, the markdown file might contain this text:

```
# EML
## Dataset
### Title
My title
```

Which would parse to this in xml:
```
<EML>
  <dataset>
    <title>My title</title>
  </dataset>
<EML>
```

To accomplish this conversion, we have built a set of parsers that convert
data in this sequence:

```
.md file -> character -> tibble -> list -> xml -> .xml file
```

Rather than export all these individual parsers, `paperbark` provides a set of
transformers with the `as_eml_` prefix, e.g.:

```{r}
library(paperbark)

c("# Heading 1", 
  "## Heading 2",
  "text content") |>
  as_eml_tibble()

c("# Heading 1", 
  "## Heading 2",
  "text content") |>
  as_eml_xml()
```

Every permutation of `as_eml_` functions is provided in `paperbark`, meaning you can 
simply choose the outcome you are looking for, and let `paperbark` handle the 
parsing chain needed to return the object class you require.

## Reading and writing

As is standard for ALA R packages, `read_` functions in `paperbark` return a `tibble`.
There are two variants, one for reading data from markdown:

```{r}
file <- system.file("example_data", 
                    "README_md_example.md", 
                    package = "paperbark")
read_md(file)
```

And a second for reading data from EML:

```{r}
read_eml("https://collections.ala.org.au/ws/eml/dr368")
```

The corresponding write functions are `write_md()` and `write_eml()`. Both
functions accept objects of all classes supported by `paperbark`, so it is possible 
to pipe read and write functions together without any intermediate stages:

```{r, eval=FALSE}
system.file("example_data", 
            "README_md_example.md", 
            package = "paperbark") |>
  read_md() |>
  write_eml("output.xml")
```

It is worth noting, though, that these are quite simple wrapper functions to 
existing import and transformation functions. For example, the code block above
can be rewritten using only one `paperbark` function, as follows:

```{r, eval=FALSE}
library(xml2)

system.file("example_data", 
            "README_md_example.md", 
            package = "paperbark") |>
  base::readLines() |>
  paperbark::as_eml_xml() |>
  xml2::write_xml("output.xml")
```
